Structural Divergence Analysis

### 2017-01-08 analysis-structure.Rmd
cambios 08/01/2017

---
title: "Structural Divergence Analysis"
author: "María Laura Marcos, Julian Echave"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, autodep = T)
```

```{r load_libraries}
# Load packages
library(ggplot2)
library(reshape2)
library(plyr)
library(quantreg)
library(bio3d)

# Source functions
source(file.path(data.dir, "multiplot.R"))
source(file.path(data.dir, "my-functions.R")) 
```

```{r set_my_colors, eval = T}
myColors = c("blue", "darkgreen", "red", "red") 
names(myColors) <- c("exp", "mut", "strong", "mut+sel")
```

@@

```{r set_run_variables}
# significance factors
f95 = 1.96
f99 = 2.6
f999 = 3.3
```

```{r set_file_names}
# input and pre-processing of data

## set up input filenames
reference.fname = file.path(data.dir, paste(family, "_ref.txt", sep = ""))
protein_list.fname = file.path(data.dir, paste(family, "_list.txt", sep = ""))
site.info.fname = file.path(data.dir, paste(p.ref, "_m.da.ca.csv", sep = ""))

dri2.exp.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.exp.norm.dr.squarei.csv", sep = ""))
Pn.exp.fname = file.path(data.dir, paste(family,"_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.exp.Pn.csv", sep = ""))
energy.exp.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.exp.va.csv", sep = ""))

dri2.strong.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.theo.norm.dr.squarei.csv", sep = ""))
Pn.strong.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.theo.Pn.csv", sep = ""))
energy.strong.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.theo.va.csv", sep = ""))

dri2.mut.fname = file.path(data.dir,paste(family, "_R0_", R0, "_beta_no.sel_K.analysis_Keff_out_m.theo.norm.dr.squarei.csv", sep = ""))
Pn.mut.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_no.sel_K.analysis_Keff_out_m.theo.Pn.csv", sep = ""))
energy.mut.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_no.sel_K.analysis_Keff_out_m.theo.va.csv", sep = ""))
```

```{r set_reference_protein}
# which is the reference protein?
reference = as.character(read.table(reference.fname, header = F)[1, 1])
```

@@

```{r set_experimental_protein_names}
# experimental protein names
protein.exp = read.table(protein_list.fname, header = F, stringsAsFactors = F)[, 1]
protein.exp = protein.exp[protein.exp != reference]
print(protein.exp)
```

[//]: Prepare site.info data frame

```{r set_up_site_info, eval = T}
# read site information of reference protein
site.info = read.csv(site.info.fname, header = T, sep = ",")

@@

# define shells of active site neighborhood (first shell is from 2.5 to 7.5, second from 7.5 to 12.5, etc)
d$shell = cut(d$da, breaks = c(-.1, 2.5, 7.5, 12.5, 17.5, 22.5, 27.5, 32.5, 37.5, 42.5, 47.5, 52.5, max(d$da)))
levels(d$shell) = seq(length(levels(d$shell))) - 1 # rename shell levels
site.info = merge(site.info, d, by = "site")

@@

[//]: Prepare experimental data

```{r read_experimental_data}
# read experimental data
dri2.exp = read.csv(dri2.exp.fname, header = T)
site = site.info$site
dri2.exp = dri2.exp[, site]
rownames(dri2.exp) = protein.exp

Pn.exp = read.csv(Pn.exp.fname, header = T)
rownames(Pn.exp) = protein.exp

energy.exp = read.csv(energy.exp.fname, header = T)
rownames(energy.exp) = protein.exp
```

[//]: Prepare data for no selection (mut)

```{r read_and_set_data_mut}
# read simulated data
dri2.mut = read.csv(dri2.mut.fname, header = T)
site = site.info$site
dri2.mut = dri2.mut[, site]
rownames(dri2.mut) = paste("mut", rownames(dri2.mut), sep = "")

Pn.mut = read.csv(Pn.mut.fname, header = T)
rownames(Pn.mut) = paste("mut", rownames(Pn.mut), sep = "")

energy.mut = read.csv(energy.mut.fname, header = T)
rownames(energy.mut) = paste("mut", rownames(energy.mut), sep = "")
```

[//]: Prepare data for strong selection

```{r read_and_set_data_strong}
# read simulated data
dri2.strong = read.csv(dri2.strong.fname, header = T)
site = site.info$site
dri2.strong = dri2.strong[, site]
rownames(dri2.strong) = paste("strong", rownames(dri2.strong), sep = "")

Pn.strong = read.csv(Pn.strong.fname, header = T)
rownames(Pn.strong) = paste("strong", rownames(Pn.strong), sep = "")

energy.strong = read.csv(energy.strong.fname, header = T)
rownames(energy.strong) = paste("strong", rownames(energy.strong), sep = "")
```

@@

Here we study how structural divergence varies from site to site. We will compare simulated data and empirical data.

```{r prepare-site-data}
# prepare my data.frame
# note: rows of dri2 are proteins, columns are sites

tdri2.strong = as.data.frame(t(as.matrix(dri2.strong)))
tdri2.strong = cbind(site.info, tdri2.strong)
siteComparison.strong = melt(tdri2.strong, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

tdri2.mut = as.data.frame(t(as.matrix(dri2.mut)))
tdri2.mut = cbind(site.info, tdri2.mut)
siteComparison.mut = melt(tdri2.mut, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

tdri2.exp = as.data.frame(t(as.matrix(dri2.exp)))
tdri2.exp = cbind(site.info, tdri2.exp)
siteComparison.exp = melt(tdri2.exp, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

siteComparison.all.datasets = rbind(
  data.frame("dataset" = "exp", siteComparison.exp),
  data.frame("dataset" = "mut", siteComparison.mut),
  data.frame("dataset" = "strong", siteComparison.strong))

@@

To measure structural divergence we use z.RSD profiles, which are the best, at least for globins (most similar distributions and profiles w.r.t experimental).

The following plot compares the mean z.RSD profiles, including standard errors (with .999 confidence level for the mean)
*Warning: the effective number of data should be smaller than the actual data, especially for the experimental dataset because of several factors, including smoothing of the profiles, phylogenetic relationships, etc. We need to deal with this for quantitative analysis. Probably we'll need to use bootstrapping to calculate standard errors non-parametrically*

```{r mean-z-RSD-vs-site}
# comparison of z.RSD profiles annotated with functional information (reported active sites).
dat = siteComparison.all.datasets
dat$dataset = revalue(dat$dataset, c("strong" = "mut+sel"))

## dat.with.means and standard errors
d = ddply(dat,c("dataset","site","AA","shell"), function(x) {
                mean.z.RSD = mean(x$z.RSD, na.rm = T)
                ndata = sum(!is.na(x$z.RSD))
                se.z.RSD = sd(x$z.RSD, na.rm = T) / sqrt(ndata)
                data.frame(mean.z.RSD, ndata, se.z.RSD)
})

## read information about minimal active site distances
m.min.da.CM = read.csv(file.path(data.dir, paste(p.ref, "_min.da.CM.ca.csv", sep = "")))

z.RSD.exp = d$mean.z.RSD[d$dataset == "exp"]
z.RSD.strong = d$mean.z.RSD[d$dataset == "mut+sel"]
z.RSD.mut = d$mean.z.RSD[d$dataset == "mut"]

plot(y = z.RSD.exp - z.RSD.strong, x = m.min.da.CM$min.da.ca, xlab = "minimal distance to active site (between CAs)")
plot(y = z.RSD.exp - z.RSD.mut, x = m.min.da.CM$min.da.CM, xlab = "minimal distance to active site (between CMs)")

print("agregar correlacion")

#d$dataset = factor(d$dataset, levels = c("exp", "mut", "strong"))
d$label = as.character(d$site)
active.sites = d$site[d$shell == 0]

p = ggplot(d,aes(x = site, y = mean.z.RSD, col = dataset, fill = dataset)) +
    geom_line() +
    geom_ribbon(aes(ymin = mean.z.RSD - f95 * se.z.RSD, ymax = mean.z.RSD + se.z.RSD), alpha = .4, show.legend = F) +
    geom_vline(xintercept = active.sites, col = "grey", show.legend = F) +
    labs(x = "site", y = "structural divergence(mean z.RSD)") +
    coord_cartesian(ylim = c(-1.5, 2.8)) +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name = NULL, values = myColors)

## plot separate and comparative plots
p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "none")

@@

#p + theme(legend.position = "bottom", legend.title = element_blank())

p + theme(legend.position = "bottom")

```

@@

```{r}
dat = siteComparison.all.datasets

# dat.with.means and standard errors
dat = ddply(dat,c("dataset", "site", "AA", "shell"), function(x) {
                mean.z.RSD = mean(x$z.RSD, na.rm = T)
                ndata = sum(!is.na(x$z.RSD))
                se.z.RSD = sd(x$z.RSD, na.rm = T) / sqrt(ndata)
                data.frame(mean.z.RSD, ndata, se.z.RSD)
})

dat.exp = subset(dat, dataset == "exp")
dat.mut = subset(dat, dataset == "mut")
dat.strong = subset(dat, dataset == "strong")

ggplot(dat, aes(x = site, y = mean.z.RSD, col = dataset)) + geom_line() +
    geom_text(data = subset(dat, dataset == "exp" & mean.z.RSD < -.5), aes(x = site, y = mean.z.RSD, label = site), nudge_y = -.1, size = 1.8)

d = dat.exp
d$d.exp.mut = dat.exp$mean.z.RSD - dat.mut$mean.z.RSD
d$d.exp.strong = dat.exp$mean.z.RSD - dat.strong$mean.z.RSD
d$d.strong.mut = dat.strong$mean.z.RSD - dat.mut$mean.z.RSD
dd = melt(data = d[, c("site", "se.z.RSD", "d.exp.mut", "d.strong.mut")], c("site", "se.z.RSD"))

ggplot(dd, aes(x = site, y = value, col = variable)) +
    geom_line() 
r = round(cor(d$d.strong.mut, d$d.exp.mut), 2)

ggplot(d, aes(x = scale(d.strong.mut), y = scale(d.exp.mut))) + 
    geom_point(aes(col = shell == 0)) +
    geom_smooth(method = "lm") +
    geom_label(aes(x = 0, y = 3, label = paste("R =", as.character(r)))) +
    labs(x = "predicted selection vs mutation (RSD change)",y="empirical vs mutation (RSD change)")

r.exp.mut = cor(dat.exp$mean.z.RSD, dat.mut$mean.z.RSD)
r.exp.strong = cor(dat.exp$mean.z.RSD, dat.strong$mean.z.RSD)
r.mut.strong = cor(dat.mut$mean.z.RSD, dat.strong$mean.z.RSD)
cor.df = data.frame("comparison" = c("exp vs mut", "exp vs strong", "mut vs strong"),
                    "R" = c(r.exp.mut, r.exp.strong, r.mut.strong))
knitr::kable(cor.df, digits = 2)
```

```{r write-pdb-files}
# write pdb files with mean.z.RSDi as b-factor for building images

## read pdb CAs
pdb.xyz.ca = ReadCA(reference.pdb.fname, chain.p.ref)$xyz

## experimental
exp.mean.z.RSD.vec <- vec2resno(as.vector(dat.exp$mean.z.RSD), sort(rep(1:ncol(pdb.xyz.ca),3)))
write.pdb(xyz = c(pdb.xyz.ca), b = exp.mean.z.RSD.vec, file =  file.path(data.dir, paste(p.ref, "_exp.pdb", sep = "")))

## mut
mut.mean.z.RSD.vec <- vec2resno(as.vector(dat.mut$mean.z.RSD), sort(rep(1:ncol(pdb.xyz.ca),3)))
write.pdb(xyz = c(pdb.xyz.ca), b = mut.mean.z.RSD.vec, file = file.path(data.dir, paste(p.ref, "_mut.pdb", sep = "")))

## mut + sel / strong
strong.mean.z.RSD.vec <- vec2resno(as.vector(dat.strong$mean.z.RSD), sort(rep(1:ncol(pdb.xyz.ca),3)))
write.pdb(xyz = c(pdb.xyz.ca), b = strong.mean.z.RSD.vec, file = file.path(data.dir, paste(p.ref, "_strong.pdb", sep = "")))
```

# Structural divergence along normal modes
In this section we consider structural divergence using normal mode coordinates.
Pn is the square divergence along a normal mode relative to the mean.

[//]: Prepare mode data

```{r prepare-mode-data}
mode = seq(ncol(Pn.strong))

# prepare modeComparison.strong
tPn.strong = as.data.frame(t(as.matrix(Pn.strong)))
tPn.strong = cbind(data.frame("mode" = mode), tPn.strong)
Pn.strong.long = melt(tPn.strong, "mode", variable.name = "protein", value.name = "Pn")

tenergy.strong = as.data.frame(t(as.matrix(energy.strong)))
tenergy.strong = cbind(data.frame("mode" = mode), tenergy.strong)
energy.strong.long = melt(tenergy.strong, "mode", variable.name = "protein", value.name = "energy")
energy.strong.long$sn2 = 1/energy.strong.long$energy

modeComparison.strong = merge(energy.strong.long, Pn.strong.long, by = c("mode", "protein"))

modeComparison.strong = ddply(modeComparison.strong, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# prepare modeComparison.mut
tPn.mut = as.data.frame(t(as.matrix(Pn.mut)))
tPn.mut = cbind(data.frame("mode" = mode), tPn.mut)
Pn.mut.long = melt(tPn.mut, "mode", variable.name = "protein", value.name = "Pn")

tenergy.mut = as.data.frame(t(as.matrix(energy.mut)))
tenergy.mut = cbind(data.frame("mode" = mode), tenergy.mut)
energy.mut.long = melt(tenergy.mut, "mode", variable.name = "protein", value.name = "energy")
energy.mut.long$sn2 = 1/energy.mut.long$energy

modeComparison.mut = merge(energy.mut.long, Pn.mut.long, by = c("mode", "protein"))
modeComparison.mut = ddply(modeComparison.mut, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# prepare modeComparison.exp
tPn.exp = as.data.frame(t(as.matrix(Pn.exp)))
tPn.exp = cbind(data.frame("mode" = mode), tPn.exp)
Pn.exp.long = melt(tPn.exp, "mode", variable.name = "protein", value.name = "Pn")

tenergy.exp = as.data.frame(t(as.matrix(energy.exp)))
tenergy.exp = cbind(data.frame("mode" = mode), tenergy.exp)
energy.exp.long = melt(tenergy.exp, "mode", variable.name = "protein", value.name = "energy")
energy.exp.long$sn2 = 1/energy.exp.long$energy

modeComparison.exp = merge(energy.exp.long,Pn.exp.long,by=c("mode", "protein"))

# renormalize Pn
modeComparison.exp = ddply(modeComparison.exp, "protein", mutate, "Pn" = Pn / mean(Pn, na.rm = T))

# modeComparison.exp has NA's because for proteins with gaps when using Keff there're less modes
# print("Warning: modeComparison <- na.omit(modeComparison)")

modeComparison.exp <- na.omit(modeComparison.exp)
modeComparison.strong <- na.omit(modeComparison.strong)
modeComparison.mut <- na.omit(modeComparison.mut)
```

[//]: Add quantile regression data

```{r}
# quantile regression

## fit median and quantiles to simulated data and add them to data.frames
fit.strong.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.strong, method = "fn")
fit.strong.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.strong, method = "fn")
fit.strong.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.strong, method = "fn")

modeComparison.strong$rq.5 = fit.strong.rq.5$fitted.values
modeComparison.strong$rq.05 = fit.strong.rq.05$fitted.values
modeComparison.strong$rq.95 = fit.strong.rq.95$fitted.values

## fit median and quantiles to simulated data and add them to data.frames
fit.mut.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.mut, method = "fn")
fit.mut.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.mut, method = "fn")
fit.mut.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.mut, method = "fn")

modeComparison.mut$rq.5 = fit.mut.rq.5$fitted.values
modeComparison.mut$rq.05 = fit.mut.rq.05$fitted.values
modeComparison.mut$rq.95 = fit.mut.rq.95$fitted.values

## fit median and quantiles to experimental data
fit.exp.rq.5 <- rq(Pn ~ sn2, tau = .5, data = modeComparison.exp, method = "fn")
fit.exp.rq.05 <- rq(Pn ~ sn2, tau = .05, data = modeComparison.exp, method = "fn")
fit.exp.rq.95 <- rq(Pn ~ sn2, tau = .95, data = modeComparison.exp, method = "fn")

modeComparison.exp$rq.5 = fit.exp.rq.5$fitted.values
modeComparison.exp$rq.05 = fit.exp.rq.05$fitted.values
modeComparison.exp$rq.95 = fit.exp.rq.95$fitted.values

## put simulated and experimental data together

modeComparison.all.datasets = rbind(data.frame("dataset" = "exp", modeComparison.exp),
                                    data.frame("dataset" = "mut", modeComparison.mut),
                                    data.frame("dataset" = "strong", modeComparison.strong))

modeComparison.all.datasets$dataset = factor(modeComparison.all.datasets$dataset, levels = c("exp", "mut", "strong"))
```

### Pn vs. mode energy

Note: lines are the median, 0.05 and 0.95 quantiles fit using rq(Pn ~ sn2).
sn2 = 1/energy is the expected MSD of along mode n. 
The mean value of Pn should be a linear function of sn2 (Pn is normalized, sn2 is not).

Here we show the square deviation along each normal mode (normalized so that <Pn> = 1) vs. the normal-mode's energy. For the empirical data, we use the actual energy obtained from Keff for the given comparison. It is this energy the one that should be used, since it's inverse is the expected value of Pn. Lines are the .5 (median), .05, and .95 quantiles obtained using `rq()`
- Upper panel: simulated vs empirical data, shown separately.
- Lower panel: simulated vs empirical overlaped and zoomed to better see the pattern.

```{r Pn-vs-mode-energy}
# plot Pn vs energy
dat = modeComparison.all.datasets
dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# p is just the median and quantiles
p.points.rq = ggplot(dat, aes(x = energy, y = Pn, col = dataset)) + 
    geom_point(alpha = .4, size = .2) +
    geom_line(aes(y = rq.5)) +
    geom_line(aes(y = rq.05)) +
    geom_line(aes(y = rq.95)) 
p = p.points.rq +  
    labs(x = "mode energy", y = "structural divergence") +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name = NULL, values = myColors)

p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "none")
p.facet %+% subset(dat, dataset == "exp")
p.facet %+% subset(dat, dataset == "mut")
p.facet %+% subset(dat, dataset == "mut+sel")
p.facet
```

### median(Pn) vs mode energy

```{r median-Pn-vs-mode-energy}
dat = modeComparison.all.datasets
dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# calculate dataset of medians

dat = ddply(dat,c("dataset", "mode"), function(x) {
    energy.mean = mean(x$energy)
    Pn.median = median(x$Pn)
    Pn.median.ci = sort(x$Pn)[qbinom(c(.025, .975), length(x$Pn), 0.5)]
    Pn.median.min = Pn.median.ci[1]
    Pn.median.max = Pn.median.ci[2]
    rq.5 = mean(x$rq.5)
    data.frame(energy.mean, Pn.median.min, Pn.median.max, Pn.median, rq.5)
})

p = ggplot(dat,aes(x=energy.mean,y=Pn.median,col=dataset,fill=dataset)) +
    geom_pointrange(aes(ymin = Pn.median.min,ymax = Pn.median.max), size = .2, alpha = .4) +
    geom_line(aes(y = rq.5))  +
    labs(x = "mode energy", y = "structural divergence (median Pn) ") +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name = NULL, values = myColors)

p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "none")
p.facet %+% subset(dat,dataset == "exp")
p.facet %+% subset(dat,dataset == "mut")
p.facet %+% subset(dat,dataset == "mut+sel")
p.facet
p + theme(legend.position = "bottom")

```

### Correlation between mode profiles

```{r}
dat = modeComparison.all.datasets

#dat$dataset = revalue(dat$dataset,c("strong" = "mut+sel"))

# calculate dataset of medians
dat = ddply(dat, c("dataset", "mode"), function(x) {
    energy.mean = mean(x$energy)
    Pn.median = median(x$Pn)
    data.frame(energy.mean, Pn.median)
})

dat = subset(dat, mode < 101)
dat.exp = subset(dat, dataset == "exp")
dat.mut = subset(dat,dataset == "mut")
dat.strong = subset(dat,dataset == "strong")

r.exp.mut = cor(dat.exp$Pn.median, dat.mut$Pn.median)
r.exp.strong = cor(dat.exp$Pn.median, dat.strong$Pn.median)
r.mut.strong = cor(dat.mut$Pn.median, dat.strong$Pn.median)
cor.df = data.frame("comparison" = c("exp vs mut", "exp vs strong", "mut vs strong"),
                    "R" = c(r.exp.mut, r.exp.strong, r.mut.strong))
knitr::kable(cor.df, digits = 2)
```

>>>>

---
title: "Structural Divergence Analysis"
author: "Maria Laura Marcos, Julian Echave"
output: html_document
---

@@

```{r setup, include = FALSE}
# set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load_packages}
# load packages
library(ggplot2)
library(reshape2)
library(plyr)
library(quantreg)
library(bio3d)

# load functions
source(file.path(data.dir, "multiplot.R"))
source(file.path(data.dir, "my-functions.R")) 
```

```{r set_my_colors}
# set my colors
myColors = c("blue", "darkgreen", "red", "red") 
names(myColors) <- c("exp", "mut", "strong", "mut+sel")
```

```{r set_run_variables}
# set significance factors
f95 = 1.96
f99 = 2.6
f999 = 3.3
```

```{r set_file_names}
# set input filenames
## data
reference.fname = file.path(data.dir, paste(family, "_ref.txt", sep = ""))
protein_list.fname = file.path(data.dir, paste(family, "_list.txt", sep = ""))
site.info.fname = file.path(data.dir, paste(p.ref, "_m.da.ca.csv", sep = ""))
min.da.CM.CA.fname = file.path(data.dir, paste(p.ref, "_min.da.CM.ca.csv", sep = ""))

## profiles to compare
dri2.exp.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_no.sel_K.analysis_Keff_out_m.exp.norm.dr.squarei.csv", sep = ""))
dri2.strong.fname = file.path(data.dir, paste(family, "_R0_", R0, "_beta_strong.sel_K.analysis_Keff_out_m.theo.norm.dr.squarei.csv", sep = ""))
dri2.mut.fname = file.path(data.dir,paste(family, "_R0_", R0, "_beta_no.sel_K.analysis_Keff_out_m.theo.norm.dr.squarei.csv", sep = ""))
```

```{r set_reference_protein}
# which is the reference protein?
reference = as.character(read.table(reference.fname, header = F)[1, 1])
if (reference != p.ref) {
  print("Warning: different references")
}
```

@@

```{r set_experimental_protein_names, comment = ""}
# experimental protein names
protein.exp = read.table(protein_list.fname, header = F, stringsAsFactors = F)[, 1]
protein.exp = protein.exp[protein.exp != reference]
print(protein.exp)
```

```{r set_site_info}
# read site information of the reference protein
site.info = read.csv(site.info.fname, header = T, sep = ",")
m.min.da.CM.CA = read.csv(min.da.CM.CA.fname)

@@

# define shells of active site neighborhood
d$shell = cut(d$da, breaks = c(-.1, 2.5, 7.5, 12.5, 17.5, 22.5, 27.5, 32.5, 37.5, 42.5, 47.5, 52.5, max(d$da)))
levels(d$shell) = seq(length(levels(d$shell))) - 1 # rename shell levels
site.info = merge(site.info, d, by = "site")

@@

```{r read_data}
# read data
## exp
dri2.exp = read.csv(dri2.exp.fname, header = T)
site = site.info$site
dri2.exp = dri2.exp[1:length(protein.exp), site]
rownames(dri2.exp) = protein.exp

## mut
dri2.mut = read.csv(dri2.mut.fname, header = T)
site = site.info$site
dri2.mut = dri2.mut[1:length(protein.exp), site]
rownames(dri2.mut) = paste("mut", rownames(dri2.mut), sep = "")

## strong
dri2.strong = read.csv(dri2.strong.fname, header = T)
site = site.info$site
dri2.strong = dri2.strong[1:length(protein.exp), site]
rownames(dri2.strong) = paste("strong", rownames(dri2.strong), sep = "")
```

@@

Here we study how structural divergence varies from site to site. We compare simulated data with empirical data.

```{r prepare_site_data}
# prepare the data
## exp
tdri2.exp = as.data.frame(t(as.matrix(dri2.exp)))
tdri2.exp = cbind(site.info, tdri2.exp)
siteComparison.exp = melt(tdri2.exp, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

## mut
tdri2.mut = as.data.frame(t(as.matrix(dri2.mut)))
tdri2.mut = cbind(site.info, tdri2.mut)
siteComparison.mut = melt(tdri2.mut, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

## strong
tdri2.strong = as.data.frame(t(as.matrix(dri2.strong)))
tdri2.strong = cbind(site.info, tdri2.strong)
siteComparison.strong = melt(tdri2.strong, id.vars = names(site.info), variable.name = "protein", value.name = "dri2")

# bluild the data.frame
siteComparison.all.datasets = rbind(data.frame("dataset" = "exp", siteComparison.exp),
                                    data.frame("dataset" = "mut", siteComparison.mut),
                                    data.frame("dataset" = "strong", siteComparison.strong))

@@

To measure structural divergence we use z.RSD profiles, which are the best for most families.

The following plot compares the mean z.RSD profiles, including standard errors (with .999 confidence level for the mean)
*Warning: the effective number of data should be smaller than the actual data, especially for the experimental dataset because of several factors. We need to deal with this for quantitative analysis. Probably we'll need to use bootstrapping to calculate standard errors non-parametrically*

```{r-mean_se_z.RSD_vs_site, fig.cap = "Black lines are the active sites and grey lines are their contacts"}
# comparison of z.RSD profiles with functional sites
## prepare dat
dat = siteComparison.all.datasets
dat$dataset = revalue(dat$dataset, c("strong" = "mut+sel"))

## dat with means and standard errors
d = ddply(dat,c("dataset", "site", "AA", "shell"), function(x) {
                mean.z.RSD = mean(x$z.RSD, na.rm = T)
                ndata = sum(!is.na(x$z.RSD))
                se.z.RSD = sd(x$z.RSD, na.rm = T) / sqrt(ndata)
                data.frame(mean.z.RSD, ndata, se.z.RSD)
})

## find the active site and its contacts
active.sites = d$site[d$shell == 0]
contacts.active.sites = d$site[d$shell == 1]

## prepare plots with mean profiles and deviations
p = ggplot(d, aes(x = site, y = mean.z.RSD, col = dataset, fill = dataset)) +
    geom_line() +
    geom_ribbon(aes(ymin = mean.z.RSD - f95 * se.z.RSD, ymax = mean.z.RSD + se.z.RSD), alpha = .4, show.legend = F) +
    geom_vline(xintercept = active.sites, show.legend = F) +
    geom_vline(xintercept = contacts.active.sites, col = "grey", show.legend = F) +
    labs(x = "site", y = "structural divergence(mean z.RSD)") +
    coord_cartesian(ylim = c(-1.5, 2.8)) +
    scale_colour_manual(name = NULL, values = myColors) + 
    scale_fill_manual(name = NULL, values = myColors)

## plot separate and comparative plots
p.facet = p + facet_grid(dataset~., switch = "y") + theme(legend.position = "bottom")

@@

## plot superimposed plots
p + theme(legend.position = "bottom")
```

@@

```{r-comparison_mean_z.RSD_vs_site}

# comparison of z.RSD profiles

ggplot(d, aes(x = site, y = mean.z.RSD, col = dataset)) +
  geom_line() +
  geom_text(data = subset(d, dataset == "exp" & mean.z.RSD < -.5), aes(x = site, y = mean.z.RSD, label = site), nudge_y = -.1, size = 1.8)+
  geom_vline(xintercept = active.sites, show.legend = F) +
  geom_vline(xintercept = contacts.active.sites, col = "grey", show.legend = F) + 
  scale_colour_manual(name = NULL, values = myColors)

# calculate correlations berween profiles
d.exp = subset(d, dataset == "exp")
d.mut = subset(d, dataset == "mut")
d.strong = subset(d, dataset == "mut+sel")

# calculate cors
r.exp.mut = cor(d.exp$mean.z.RSD, d.mut$mean.z.RSD)
r.exp.strong = cor(d.exp$mean.z.RSD, d.strong$mean.z.RSD)
r.mut.strong = cor(d.mut$mean.z.RSD, d.strong$mean.z.RSD)

cor.df = data.frame("comparison" = c("exp vs mut", "exp vs strong", "mut vs strong"),
                    "R" = c(r.exp.mut, r.exp.strong, r.mut.strong))

knitr::kable(cor.df, digits = 2, caption = "Correlation between <Z-scores> profiles")
```

### Distances to the active sites

```{r-comparison-da}

# comparison of means.z.scores with the minimum distance to the functional sites
plot(y = d.exp$mean.z.RSD - d.mut$mean.z.RSD, x = m.min.da.CM.CA$min.da.ca, xlab = "minimal distance to active site (between CAs)", ylab = "exp.mean.z.RSD - mut.mean.z.RSD")
plot(y = d.exp$mean.z.RSD - d.mut$mean.z.RSD, x = m.min.da.CM.CA$min.da.CM, xlab = "minimal distance to active site (between CMs)", ylab = "exp.mean.z.RSD - mut.mean.z.RSD")

## calculate correlations between profiles
r.min.da.CA = cor((d.exp$mean.z.RSD - d.mut$mean.z.RSD), m.min.da.CM.CA$min.da.ca)
r.min.da.CM = cor((d.exp$mean.z.RSD - d.mut$mean.z.RSD), m.min.da.CM.CA$min.da.CM)

cor.df = data.frame("comparison" = c("(exp - mut) vs min.da.CA", "(exp - mut) vs min.da.CM"), "R" = c(r.min.da.CA, r.min.da.CM))

knitr::kable(cor.df, digits = 2, caption = "Correlation between differences in <Z-scores> profiles and the minimal distance to the active site")
```

```{r-comparisons-between-plots}

# plot exp vs simulated profiles 
par(pty = "s")

plot(x = d.exp$mean.z.RSD, 
     y = d.mut$mean.z.RSD, 
     xlab = "exp.mean.z.RSD", 
     ylab = "mut.mean.z.RSD", 
     xlim = c(min(c(d.exp$mean.z.RSD, d.mut$mean.z.RSD)), max(c(d.exp$mean.z.RSD, d.mut$mean.z.RSD))),
     ylim = c(min(c(d.exp$mean.z.RSD, d.mut$mean.z.RSD)), max(c(d.exp$mean.z.RSD, d.mut$mean.z.RSD))))
points(x = as.numeric(d.exp$mean.z.RSD)[active.sites], y = as.numeric(d.mut$mean.z.RSD)[active.sites], col = "yellow")
points(x = as.numeric(d.exp$mean.z.RSD)[contacts.active.sites], y = as.numeric(d.mut$mean.z.RSD)[contacts.active.sites], col = "green")
abline(0, 1)

plot(x = d.exp$mean.z.RSD, 
     y = d.strong$mean.z.RSD, 
     xlab = "exp.mean.z.RSD", 
     ylab = "strong.mean.z.RSD", 
     xlim = c(min(c(d.exp$mean.z.RSD, d.strong$mean.z.RSD)), max(c(d.exp$mean.z.RSD, d.strong$mean.z.RSD))),
     ylim = c(min(c(d.exp$mean.z.RSD, d.strong$mean.z.RSD)), max(c(d.exp$mean.z.RSD, d.strong$mean.z.RSD))))
points(x = as.numeric(d.exp$mean.z.RSD)[contacts.active.sites], y = as.numeric(d.strong$mean.z.RSD)[contacts.active.sites], col = "green")
points(x = as.numeric(d.exp$mean.z.RSD)[active.sites], y = as.numeric(d.strong$mean.z.RSD)[active.sites], col = "yellow")
abline(0, 1)

plot(x = d.mut$mean.z.RSD, 
     y = d.strong$mean.z.RSD, 
     xlab = "mut.mean.z.RSD", 
     ylab = "strong.mean.z.RSD", 
     xlim = c(min(c(d.mut$mean.z.RSD, d.strong$mean.z.RSD)), max(c(d.mut$mean.z.RSD, d.strong$mean.z.RSD))),
     ylim = c(min(c(d.mut$mean.z.RSD, d.strong$mean.z.RSD)), max(c(d.mut$mean.z.RSD, d.strong$mean.z.RSD))))
points(x = as.numeric(d.mut$mean.z.RSD)[contacts.active.sites], y = as.numeric(d.strong$mean.z.RSD)[contacts.active.sites], col = "green")
points(x = as.numeric(d.mut$mean.z.RSD)[active.sites], y = as.numeric(d.strong$mean.z.RSD)[active.sites], col = "yellow")
abline(0, 1)
```
